---
title: "Differential expression"
output: html_document
---

Differential expression between control samples and test/stimulated samples. Wald test or likelihood ratio test (LRT) test used to analyse differential expression between conditions. 

```{r, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(cache=TRUE, warning=FALSE, message=FALSE)
source('functions.R')
library(org.Hs.eg.db)
library(DESeq2)
library(pheatmap)
library(tidyverse)
library(DT)
library(yaml)
library(ggplot2) 
library(ggrepel) 
```

```{r yaml, echo=TRUE,warning=FALSE,message=FALSE,error=FALSE, include=FALSE}
params <- read_yaml("../pipeline.yml")
```

```{r imports, echo=TRUE,warning=FALSE,message=FALSE,error=FALSE, include=FALSE}

df_mRNA <- read.table(gzfile("../merged_idxstats.txt.gz"), sep = "\t", header = TRUE, row.names = 1, check.names=FALSE)

colnames(df_mRNA) <- gsub("_trna", "", x = colnames(df_mRNA), fixed = T)
meta_data <- read.table("../meta_data.csv", sep=",", header = TRUE)

rownames(meta_data) <- meta_data$Sample
df_mRNA = df_mRNA[,rownames(meta_data)]

all(rownames(meta_data) %in% colnames(df_mRNA))
```


```{r , message=FALSE, include=FALSE}
## Set up the DESeq2 object
# Decide if LRT or Wald test is to be ran
if (params$diff_eprs$model == "Wald") {
  res <- run_deseq2(as.data.frame(df_mRNA), meta_data, control=params$Wald$control, test=params$Wald$test, value=params$Wald$value, design=params$diff_eprs$design)
  de_test <- "Wald"
} else{
  res <- run_deseq2_LRT(as.data.frame(df_mRNA), meta_data, design=params$diff_eprs$design,
                        full=params$LRT$full, reduced=params$LRT$reduced)
  de_test <- "LRT"
}



```

# Results - summary
`r de_test` test used for DE analysis. This can be changed in the pipeline.yml file. 

```{r , echo=FALSE}
summary(res, alpha=0.01)

res_output<- as.data.frame(res)
write.csv(res_output, file="res.csv")
```

Download significant results (p < 0.01):
<a href="res.csv" download=""><button class="btn"><i class="fa fa-download"></i> Download</button></a>

# Results - full table

```{r, echo = FALSE}
datatable(res_output)
```


# MA plot

```{r MA, echo = FALSE}
DESeq2::plotMA(res)
```


# Volcano plot

```{r volcano, echo=FALSE}
test <- as.data.frame(res)
test <- test %>% 
  rownames_to_column("SYMBOL")

mutateddf <- dplyr::mutate(test, sig=ifelse(res$padj<0.01, "padj<0.01", "Not Sig")) #Will have different colors depending on significance
input <- cbind(gene=rownames(mutateddf), mutateddf )
input <- input %>% 
  arrange(input$padj)

symbol_data <- head(input, 10)

#convert the rownames to a column
volc = ggplot(input, aes(log2FoldChange, -log10(padj))) + #volcanoplot with log2Foldchange versus pvalue
    geom_point(aes(col=sig)) + #add points colored by significance
geom_point(data=symbol_data, aes(log2FoldChange, -log10(padj)), colour="red") 

volc+geom_text_repel(data=symbol_data, aes(label=`SYMBOL`)) + scale_colour_Publication() + theme_bw()#adding text for the genes


```


