---
title: "Differential expression"
output: html_document
---

Differential expression between control samples and test/stimulated samples. Wald test or likelihood ratio test (LRT) test used to analyse differential expression between conditions. 

```{r, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(cache=TRUE, warning=FALSE, message=FALSE)
source('functions.R')
library(DESeq2)
library(pheatmap)
library(tidyverse)
library(DT)
library(yaml)
library(ggplot2) 
library(ggrepel)
library(ggpubr)
```

```{r yaml, echo=TRUE,warning=FALSE,message=FALSE,error=FALSE, include=FALSE}
params <- read_yaml("../pipeline.yml")
```

```{r imports, echo=TRUE,warning=FALSE,message=FALSE,error=FALSE, include=FALSE}


design_files <- list.files(path="..",pattern = "^design_")

if (file.exists("../merged_idxstats.txt.gz")) {
  df_mRNA <- read.table(gzfile("../merged_idxstats.txt.gz"), sep = "\t", header = TRUE, row.names = 1)
  colnames(df_mRNA) <- gsub(".", "-", x = colnames(df_mRNA), fixed = T)
} else {
  print("Please add featurecounts.tsv.gz into the project folder as it seems to be missing")
}


if (file.exists(paste0("../", design_files[1]))) {
  for (i in design_files){
  meta.loc <- paste0("../", i)
  meta_data <- read.table(meta.loc, sep=",", header = TRUE) 
  rownames(meta_data) <- paste0(meta_data$Sample, "_trna")
  df_mRNA_tmp = df_mRNA[,rownames(meta_data)]
  all(rownames(meta_data) %in% colnames(df_mRNA_tmp))
  assign(paste("meta_data", i, sep = "."), meta_data)
  }
} else {
    print("No design files were detected please add a file called design_<test>_<control>_<test>_<column>.csv. Please refer to documentation on github for more ifnormation")
  }

```


```{r , message=FALSE, include=FALSE}
## Set up the DESeq2 object
## design_<test>_<control>_<test>_<column>.csv
for (i in design_files) {
  if(stringr::str_detect(i, "csv~")){
    next
  }
  x <- strsplit(i, "_")
  stat.test <- x[[1]][2]
  control <- x[[1]][3]
  test <- x[[1]][4]
  value <- x[[1]][5]
  value <- gsub(".csv","",value)
  meta_data <- get(gsub("SAMPLE_FILE",i , "meta_data.SAMPLE_FILE"))
  model <- as.character(meta_data$model[[1]])
  
  if ("wald" == stat.test){
    deseq_results <- run_deseq2(as.data.frame(df_mRNA), meta_data, control = control, test=test, value=value, model = model)

  } else {
    reduced_model <- as.character(meta_data$reduced[[1]])
      deseq_results <- run_deseq2_lrt(as.data.frame(df_mRNA), meta_data, control = control, test=test, value=value, model = model, reduced = reduced_model)
    }
  
  res = deseq_results@res
  dds = deseq_results@dds
  assign(paste("res", i, sep = "."), res)
  assign(paste("dds", i, sep = "."), dds)
}



```


# Model fitting


This section of the report describes figures that can be used to assess how well the DESeq2 model has fitted the data.

## MA plots {.tabset .tabset-fade}

In DESeq2, the function plotMA shows the log2 fold changes attributable to a given variable over the mean of normalized counts for all the samples in the DESeqDataSet. Points will be colored red if the adjusted p value is less than 0.01. Points which fall out of the window are plotted as open triangles pointing either up or down.


```{r, results='asis', echo = FALSE}
for (i in design_files){
  name <-  gsub(".csv","",i)
  res <- get(gsub("SAMPLE_FILE",i , "res.SAMPLE_FILE"))
  cat("### ",name,"\n")
  plt <- DESeq2::plotMA(res)
  print(plt)
  cat('\n\n')
} 
```

## Volcano plots {.tabset .tabset-fade}



```{r, results='asis', echo = FALSE}

for (i in design_files){
  name <-  gsub(".csv","",i)
  res <- get(gsub("SAMPLE_FILE",i , "res.SAMPLE_FILE"))
  cat("### ",name,"\n")
  plt <- plot_volcano(res)
  print(plt)
  cat('\n\n')
} 


```

## Results tables {.tabset .tabset-fade}

The results tables for the significant and full differential expression for tRNAs are output in the results/ directory of the Report.dir folder.

```{r data.setup}
new_list <- list()
names <- c()
for (i in design_files){
  name.start <-  gsub(".csv","",i)
  res <- get(gsub("SAMPLE_FILE",i , "res.SAMPLE_FILE"))
  name <- paste0(name.start)
  dt <- filter_genes(as.data.frame(res), name=name)
  table <- dt@sig
  new_list[[paste0("dt_",name,sep="")]] <- table
  names <- c(names, name)
} 
df <- data.frame(cbind(new_list, classification=names))

```


